How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-08 doc/paper.tex
moved to biber. moved more to supplement

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

@@

@@

\begin{abstract}
  One of the central tenants of conservation paleobiology is that by studying the past we can better predict the future. It has been suggested that risk assessments for extant species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. Such suggestions prompt two key questions: (1) At a given timescale, to what extent are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) At a given point in time, how accurate do extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns  turn out to be? To address these questions we analyze the exceptionally well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, climate state, and the change in climate state. Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes. AN important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

By studying how species vary in their extinction risk over time and we can assess which species are at greater risk under unobserved conditions. We know that a species' risk of extinction varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}. Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.

@@

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}. We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. This begs the question: how does a species' extinction risk change over its duration? While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}). For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years? Here, we explicitly model and quantify the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Similarly, we include species geologic age at time of observation as a potential predictor of extinction -- a factor that may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms F(oraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible.

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts. Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.

@@

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this distance was measured in kilometers. Geographic range was then log-plus-one transformed, and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval. Temperature was then transformed and standardized the in the same manner as geographic range. Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.

@@

@@

See our code repository LINK for full code details. Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.

@@

This presents an interesting conundrum about how to improve upon our results. A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved. However, if we decrease the amount of data in our model, our results by definition decrease in their quality. For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)

@@

% bibliographic information
\bibliographystyle{abbrvnat}
\bibliography{citations}

@@

>>>>

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
%\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[style=authoryear-icomp,
            backend=biber,
            bibencoding=utf8,
            natbib=true,
            url=false,
            doi=false,
            isbn=false]{biblatex}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

@@

% bibliographic information
%\bibliographystyle{abbrvnat}
%\bibliography{citations}
\addbibresource{citations.bib}

@@

\begin{abstract}
  One of the central tenants of conservation paleobiology is that by studying the past we can better predict the future. It has been suggested that risk assessments for extant species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. Such suggestions prompt two key questions: (1) At a given timescale, to what extent are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) At a given point in time, how accurate do extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns  turn out to be? To address these questions we analyze the exceptionally well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, climate state, and the change in climate state. Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. However, models based on present geographic ranges and past range-risk relationships are likely to be modernly successful at predicting future extinction/survival outcomes. AN important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
  %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

Studying how species vary in their extinction risk over time means we can assess which species are at greater risk under unobserved conditions. We know that a species' risk of extinction varies over time in both intensity (average rate) and selectivity (difference in risk between taxa) \citep{Payne2007,Payne2016,Ezard2011}. Species, after all, can go extinct at any ``moment'' and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction. What has not been evaluated is that as extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? By specifically including and modeling the temporal variation in extinction risk, we are able to improve our overall predictions because we incorporate and explicitly model differences between observations from across a range extinction intensities and selectivities.

@@

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}. We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. This begs the question: how does a species' extinction risk change over its duration? While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}). 
%For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years? 
Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms F(oraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. 
%Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My (million year) scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. 
Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

%Being able to analyze over 60 million years of fossil occurrences allows to actually quantify how accurate our predictions are in general, but also how much variation there is in predictive accuracy over time and in many different environmental contexts. Specifically, we might expect that our model's predictive performance is best during prolonged periods of similar stress, such as the Eocene-Miocene transition \citep{Zachos2008} -- more samples from similar environments inherently improves future predictions in unobserved, but similar conditions. Alternatively, we would expect our model based predictions of extinction surrounding the Paleocene-Eocene Thermal Maximum may be less accurate because there are inherently fewer samples from the rapid climatic event \citep{Zachos2008}.

@@

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. 
After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this distance was measured in kilometers.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned and how our covariates were standardized and transformed prior to analysis.

@@

Importantly, we include species geologic age at time of observation as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}. Again, see Section \ref{sec:model_desc} for how this term was specified as part of our model.

@@

See our code repository https://github.com/psmits/trident for full code details. Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.

@@

%This presents an interesting conundrum about how to improve upon our results. A simple hypothesis for how to improve upon our results is that if we were to include more biological information in our model, our estimates of species relative extinction risk would be improved. However, if we decrease the amount of data in our model, our results by definition decrease in their quality. For example, compare the results from our models fit using full in-sample dataset, and the results from the cross-validation where the fit to each fold is by definition based on less than full information.)

@@

\printbibliography

@@

\subsection{Data Specifications} \label{sec:data_desc}

\subsubsection{Binning fossil occurrences}

Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

\subsubsection{Covariate transforms and standardizations} \label{sec:data_trans}

Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.

Temperature was also transformed and standardized the in the same manner as geographic range. Difference in temperature between between observations was measured from the standardized temperature values and was not standardized separately.
